---
title: "Final Project"
output: pdf_document
extra_dependencies: ["pdflscape", "longtable"]
date: "2025-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(lubridate)
library(ggplot2)
library(scales)
library(patchwork)
```

```{r, message=FALSE, echo=FALSE}
FundData <- read_csv("FinalProject_FundData.csv") %>% select(-1) %>% 
  mutate(date = myd(paste(date, "1"), truncated = TRUE))
FundInfo <- read_csv("FinalProject_FundInfo.csv") %>% select(-1)
FundReturns <- FundData %>% select(1, 11:ncol(FundData))

FundInfo <- FundInfo %>%
  mutate(
    Equity_Style = case_when(
      grepl("V$", Style) ~ "Value",
      grepl("G$", Style) ~ "Growth",
      grepl("B$", Style) ~ "Blend"
    ),
    Market_Cap = case_when(
      grepl("^R1000", Style) ~ "Large Cap",
      grepl("^RMid", Style) ~ "Mid Cap",
      grepl("^R2000", Style) ~ "Small Cap"
    ),
    Benchmark_Ticker = case_when(
      Style == "R1000B" ~ "IWB",
      Style == "R1000V" ~ "IWD",
      Style == "R1000G" ~ "IWF",
      Style == "R2000B" ~ "IWM",
      Style == "R2000V" ~ "IWN",
      Style == "R2000G" ~ "IWO",
      Style == "RMidB"  ~ "IWR",
      Style == "RMidV"  ~ "IWS",
      Style == "RMidG"  ~ "IWP"
    )
  )
```

```{r, message=FALSE, echo=FALSE}
AggregatedWeights <- FundInfo %>%
  group_by(Market_Cap, Equity_Style) %>%
  summarise(
    weight_B = sum(weight_B),
    weight_A = sum(weight_A),
    .groups = "drop"
  )

BenchmarkReturns <- FundData %>%
  select(date, all_of(unique(FundInfo$Benchmark_Ticker)))

StyleBenchmarks <- FundInfo %>%
  select(Market_Cap, Equity_Style, Benchmark_Ticker) %>%
  distinct() %>%
  left_join(
    BenchmarkReturns %>% pivot_longer(-date, names_to = "Benchmark_Ticker", values_to = "BenchmarkReturn"),
    by = "Benchmark_Ticker"
  )

FundReturns <- FundData %>%
  pivot_longer(-date, names_to = "Ticker", values_to = "FundReturn") %>%
  left_join(FundInfo %>% select(Ticker, Market_Cap, Equity_Style, weight_A), by = "Ticker") %>% na.omit()

Portfolio_Segment_Return <- FundReturns %>%
  group_by(date, Market_Cap, Equity_Style) %>%
  summarise(
    portfolio_segment_return = sum(FundReturn * weight_A) / sum(weight_A),
    .groups = "drop"
  ) %>%
  left_join(
    StyleBenchmarks %>% select(date, Market_Cap, Equity_Style, BenchmarkReturn),
    by = c("date", "Market_Cap", "Equity_Style")
  ) %>%
  left_join(AggregatedWeights, by = c("Market_Cap", "Equity_Style"))
```

```{r, message=FALSE, echo=FALSE}
style <- Portfolio_Segment_Return %>%
  group_by(date, Equity_Style) %>%
  summarise(
    
    return_B = sum(weight_B * BenchmarkReturn) / sum(weight_B),
    
    return_A = sum(weight_A * portfolio_segment_return) / sum(weight_A),
    
    weight_B = sum(weight_B),
    weight_A = sum(weight_A),
    
    .groups = "drop"
  )

overall_benchmark <- style %>%
  group_by(date) %>%
  summarise(overall_benchmark = sum(weight_B * return_B), .groups = "drop")

style <- style %>%
  left_join(overall_benchmark, by = "date")

style <- style %>%
  mutate(
    Allocation_Effect = (weight_A - weight_B) * (return_B - overall_benchmark),
    
    Implementation_Effect = weight_B * (return_A - return_B),
    
    Interaction_Effect = (weight_A - weight_B) * (return_A - return_B),
    
    Total_Effect = Allocation_Effect + Implementation_Effect + Interaction_Effect
  )
```

```{r, message=FALSE, echo=FALSE}
size <- Portfolio_Segment_Return %>%
  group_by(date, Market_Cap) %>%
  summarise(
    
    return_B = sum(weight_B * BenchmarkReturn) / sum(weight_B),
    
    return_A = sum(weight_A * portfolio_segment_return) / sum(weight_A),
    
    weight_B = sum(weight_B),
    weight_A = sum(weight_A),
    
    .groups = "drop"
  )

overall_benchmark <- size %>%
  group_by(date) %>%
  summarise(overall_benchmark = sum(weight_B * return_B), .groups = "drop")

size <- size %>%
  left_join(overall_benchmark, by = "date")

size <- size %>%
  mutate(
    Allocation_Effect = (weight_A - weight_B) * (return_B - overall_benchmark),
    
    Implementation_Effect = weight_B * (return_A - return_B),
    
    Interaction_Effect = (weight_A - weight_B) * (return_A - return_B),
    
    Total_Effect = Allocation_Effect + Implementation_Effect + Interaction_Effect
  )
```

```{r, message=FALSE, echo=FALSE}
FundReturns <- FundData %>%
  pivot_longer(-date, names_to = "Ticker", values_to = "FundReturn") %>%
  left_join(FundInfo %>% select(Name, Ticker, Benchmark_Ticker, Market_Cap, Equity_Style, weight_B, weight_A), by = "Ticker") %>%
  left_join(BenchmarkReturns %>% pivot_longer(-date, names_to = "Benchmark_Ticker", values_to = "BenchmarkReturn"), by = c("date", "Benchmark_Ticker")) %>% na.omit()

Manager_Analysis <- FundReturns %>%
  group_by(date, Market_Cap, Equity_Style) %>%
  mutate(
    norm_weight_B = weight_B / sum(weight_B, na.rm = TRUE),
    norm_weight_A = weight_A / sum(weight_A, na.rm = TRUE),
    overall_benchmark = sum(norm_weight_B * BenchmarkReturn, na.rm = TRUE)
  ) %>%
  ungroup()

Manager_Analysis <- Manager_Analysis %>%
  mutate(
    Allocation_Effect = (norm_weight_A - norm_weight_B) * (BenchmarkReturn - overall_benchmark),
    Implementation_Effect = norm_weight_B * (FundReturn - BenchmarkReturn),
    Interaction_Effect = (norm_weight_A - norm_weight_B) * (FundReturn - BenchmarkReturn),
    Total_Effect = Allocation_Effect + Implementation_Effect + Interaction_Effect
  )
```

# Key Results of Time-Series Analysis
## Allocation vs. Implementation Effects
\textbf{Allocation effects were near zero} across all size tiers and styles, indicating a \textbf{passive allocation strategy} closely aligned with the benchmark. While this reduces tracking error, it limits opportunities for excess returns through active positioning. Conversely, \textbf{implementation effects varied}, with \textbf{Large Cap Growth showing significant underperformance (-0.0029)} due to weak stock selection, while \textbf{Large Cap Value (0.0010) and Small Cap Growth (0.0005) outperformed}, suggesting better security selection.

## Trends in Performance
\begin{itemize}
    \item \textbf{Large Cap Growth struggled (-0.003)}, likely due to valuation sensitivity or poor stock selection.
    \item \textbf{Large Cap Value outperformed (0.0009)}, benefiting from stronger fundamentals.
    \item \textbf{Mid Cap Blend showed slight positive results (0.0004)} while Growth remained neutral.
    \item \textbf{Small Cap Growth displayed potential} but lacked consistency.
\end{itemize}

## Manager Information Ratios
\textbf{Most managers had weak risk-adjusted returns}, with \textbf{IRs below 0.2}. A notable underperformer was \textbf{BlackRock Large Cap Focus Value Fund (IR = -0.183)}, despite active positioning. However, \textbf{Ave Maria Growth Fund (IR = 0.128)} and \textbf{Lord Abbett Micro Cap Growth Fund (IR = 0.120)} stood out as \textbf{top performers}, delivering superior risk-adjusted returns. 

## Recommendations
\begin{itemize}
    \item \textbf{Enhance Growth Stock Selection:} Large Cap Growth underperformance suggests weaknesses in security selection—factor-based screening should be considered.
    \item \textbf{Increase Flexibility in Value Strategies:} Value-focused managers consistently delivered positive results and should be given additional capital.
    \item \textbf{Strengthen Risk Management:} High tracking error in some funds, like Boston Partners All Cap Value Fund, suggests the need for tighter risk controls.
\end{itemize}

By addressing these weaknesses while leveraging top-performing funds, the firm can enhance overall investment performance.

# Areas for Improvement
## Growth Stock Selection and Passive Allocation Strategy
The analysis highlights two key areas for improvement: \textbf{Growth stock selection} and \textbf{passive allocation strategy}. Large Cap Growth underperformed (Implementation Effect = -0.0029), suggesting managers may be \textbf{overpaying for momentum stocks}. A \textbf{factor exposure review} is needed to integrate value and profitability filters. Additionally, near-zero allocation effects indicate a \textbf{passive management approach}, limiting \textbf{alpha opportunities}. Tactical shifts, particularly in \textbf{Value strategies}, could enhance returns.

## Risk Management Considerations
Some funds, like Boston Partners All Cap Value Fund (Tracking Error = 0.014, Active Return = -0.002), took risks but failed to deliver alpha. \textbf{Tighter active risk limits} should reduce off-benchmark bets. Growth strategies also showed \textbf{high implementation volatility}, requiring \textbf{risk-adjusted constraints} for more consistent performance. Strengthening stock selection, allocation flexibility, and risk controls will improve overall investment outcomes.

# Opportunities for Additional Flexibility
## Expanding Strategic Bets
Expanding \textbf{Value strategies} is key, as \textbf{Large Cap Value} consistently outperforms Growth. Increasing allocations to \textbf{top-performing Value funds} can enhance returns. \textbf{Small Cap Growth} also shows promise, with \textbf{Lord Abbett Micro Cap Growth Fund} delivering the best \textbf{risk-adjusted returns}, warranting expanded mandates.

## Capitalizing on Strong Performers
To capitalize on strong performers, \textbf{high Information Ratio managers} should receive larger mandates. \textbf{Ave Maria Growth Fund} and \textbf{Lord Abbett Micro Cap Growth Fund} stand out, justifying \textbf{greater capital allocations and leadership roles} in strategy development.

# Observed Patterns
\textbf{Value strategies consistently outperformed Growth}, likely due to cyclical macroeconomic trends favoring undervalued equities. \textbf{Large Cap} was dominated by \textbf{Value strategies}, while \textbf{Small Cap Growth showed potential} but was inconsistent. A small subset of managers contributed most positive results, while \textbf{Growth managers struggled across multiple tiers}.

# Conclusion
The firm's \textbf{investment process is effective in Value strategies but struggles in Growth stock selection and active allocation}. To enhance performance, I recommend the following:

\begin{enumerate}
    \item \textbf{Revamp Growth Research \& Factor Exposure:} Address implementation gaps in Growth strategies to prevent further underperformance.
    \item \textbf{Empower Top-Performing Managers:} Grant flexibility and increased allocations to \textbf{high IR managers}.
    \item \textbf{Adopt a More Dynamic Allocation Strategy:} Move from \textbf{passive benchmarking} to \textbf{tactical positioning} in areas with clear alpha-generation potential.
\end{enumerate}


\newpage

# Appendix
```{r, echo=FALSE, message=FALSE, combined_plot, fig.width=12, fig.height=6, out.width="100%"}
size_plot <- size %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = Allocation_Effect, color = "Allocation"), linewidth = 0.8) +
  geom_line(aes(y = Implementation_Effect, color = "Implementation"), linewidth = 0.8) +
  facet_wrap(~Market_Cap, ncol = 1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Size Tier Performance",
       x = "Date", y = "Effect Value", color = "Metric") +
  theme_minimal() +
  scale_color_manual(values = c("Allocation" = "#1f77b4", "Implementation" = "#ff7f0e"))

style_plot <- style %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = Allocation_Effect, color = "Allocation"), linewidth = 0.8) +
  geom_line(aes(y = Implementation_Effect, color = "Implementation"), linewidth = 0.8) +
  facet_wrap(~Equity_Style, ncol = 1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Style Decision Performance",
       x = "Date", y = "Effect Value", color = "Metric") +
  theme_minimal() +
  scale_color_manual(values = c("Allocation" = "#1f77b4", "Implementation" = "#ff7f0e"))

combined_plot <- size_plot + style_plot + 
  plot_layout(ncol = 2, guides = "collect") & theme(legend.position = "bottom")

print(combined_plot)
```


```{r, echo=FALSE}
avg_performance <- bind_rows(
  size %>%
    group_by(Market_Cap) %>%
    summarise(across(c(Allocation_Effect, Implementation_Effect, Total_Effect), mean),
              .groups = "drop") %>%
    mutate(Category = "Size"),
  
  style %>%
    group_by(Equity_Style) %>%
    summarise(across(c(Allocation_Effect, Implementation_Effect, Total_Effect), mean),
              .groups = "drop") %>%
    rename("Market_Cap" = "Equity_Style") %>%
    mutate(Category = "Style")
) %>%
  select(Category, everything())

style_cap_table <- Manager_Analysis %>%
  group_by(Market_Cap, Equity_Style) %>%
  summarise(
    Avg_Allocation = mean(Allocation_Effect),
    Avg_Implementation = mean(Implementation_Effect),
    Avg_Total = mean(Total_Effect),
    .groups = "drop"
  )

avg_performance %>%
  kable(format = "latex", booktabs = TRUE, digits = 4,
        caption = "Average Performance by Category") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)

style_cap_table %>%
  kable(format = "latex", booktabs = TRUE, digits = 4,
        caption = "Style/Cap Tier Performance") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r, echo=FALSE}
info_ratio_table <- Manager_Analysis %>%
  group_by(Name, Market_Cap, Equity_Style) %>%
  summarise(
    Active_Return = mean(Total_Effect),
    Tracking_Error = sd(Total_Effect),
    Information_Ratio = Active_Return / Tracking_Error,
    .groups = "drop"
  ) %>%
  arrange(Market_Cap, Equity_Style, desc(Information_Ratio))

info_ratio_table %>%
  kable(format = "latex", booktabs = TRUE, digits = 3,
        caption = "Manager Information Ratios") %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  row_spec(which(info_ratio_table$Information_Ratio > 0.5), bold = TRUE, background = "#e6f5d0")
```

```{r, echo=FALSE}
december_style <- style %>%
  filter(lubridate::month(date) == 12) %>%
  select(date, Equity_Style, Allocation_Effect, Implementation_Effect, Total_Effect)

print(december_style)
```

```{r, echo=FALSE}
december_size <- size %>%
  filter(lubridate::month(date) == 12) %>%
  select(date, Market_Cap, Allocation_Effect, Implementation_Effect, Total_Effect)

print(december_size)
```

```{r, echo=FALSE}
December_Effect_Table <- Manager_Analysis %>%
  filter(month(date) == 12) %>%
  select(date, Name, Market_Cap, Equity_Style, Allocation_Effect, Implementation_Effect, Total_Effect) %>%
  arrange(date, Market_Cap, Equity_Style, Name)

December_Effect_Table %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "December Effect Analysis",
    longtable = TRUE
  ) %>%
  kable_styling(
    latex_options = c("scale_down", "repeat_header"),
    font_size = 7,
    full_width = FALSE
  ) %>%
  landscape()
```

